---
title: "Disease compartment models"
output: html_document
---


```{r, echo=F, message=F}
library(deSolve) #ode solver
```

We'll start by investigating the Susceptible-Infected (SI) model and the Susceptible-Infected-Susceptible (SIS) model, the simplest version of a disease compartment model. It describes how susceptible individuals can become infected then become susceptible again. It is of the form 


$$
\frac{dS}{dt} = -\beta S I + \gamma I \\
\frac{dI}{dt} = \beta S I - \gamma I.
$$

We'll start with the SI model, a special case of the SIS model when $\gamma=0$. 

```{r setup}
library(deSolve)
# parameters
pars <- c(beta=0.001, gamma=0)

# initial state 
init <- c(S = 99, I=1)

# times to solve teh equation at
times <- seq(0, 100, by = 0.1)
```

```{r SImodel}
SIS_mod <- function(t, state, pars) {
  with(as.list(c(state, pars)), {
    d_S <- -beta*I*S + gamma*I
    d_I <-  beta*I*S - gamma*I
    return(list(c(S = d_S, I = d_I)))
  })
}

#this line runs the model
SI_model <- ode(init, times, SIS_mod, pars) 
```


Below we plot the simulated abundances over time.

```{r plotModel}
plot(SI_model[,1], SI_model[,2], type='l', lwd=2, xlab="Time", ylab="Abundance", ylim=c(0,100))
lines(SI_model[,1], SI_model[,3], col="cornflowerblue", lwd=2)
```


**Excercises**

- Mess around with value of the parameter $\beta$. What if $\beta=0$? 

- Does the initial population state affect the resulting population dynamics?


Now we will relax the assumption that the recovery rate is 0. We will assume that infected individuals will recover from infection and immediately become susceptible again.

Use the code below to answer the questions about the SIS model.


```{r SISmodel}
pars <- c(beta=0.001, gamma=0)

# initial state 
init <- c(S = 99, I=1)

# times to solve teh equation at
times <- seq(0, 100, by = 0.1)

#this line runs the model
SIS_model <- ode(init, times, SIS_mod, pars) 

plot(SIS_model[,1], SIS_model[,2], type='l', lwd=2, xlab="Time", ylab="Abundance", ylim=c(0,100))
lines(SIS_model[,1], SIS_model[,3], col="cornflowerblue", lwd=2)

```

**Excercises**

- Before you start messing with the model, think about what you expect will happen to the dynamics when infecteds can become susceptible again. Will all the individuals still become infected? 

- Try increasing the value of the $\gamma$ parameter. What does this do to the resulting dynamics? Is this what you expected including recovery would do? 



## The SIR model

Now we'll add a Recovery compartment (this could also be a Removed compartment if the infection is fatal). We'll assume that infected individuals become recovered with rate $\gamma$.


```{r SIRinit}
# parameters
pars <- c(beta=0.001, gamma=0.01)

# initial state 
init <- c(S = 99, I=1, R=0)

# times to solve the equation at
times <- seq(0, 100, by = 0.1)
```

```{r SIRmodel}
SIR_mod <- function(t, state, pars) {
  with(as.list(c(state, pars)), {
    d_S <- -beta*I*S 
    d_I <-  beta*I*S - gamma*I
    d_R <-  gamma*I
    return(list(c(S=d_S, I=d_I, R=d_R)))
  })
}

#this line runs the model
SIR_model <- ode(init, times, SIR_mod, pars) 

plot(SIR_model[,1], SIR_model[,2], type='l', lwd=2, xlab="Time", ylab="Abundance", ylim=c(0,100))
lines(SIR_model[,1], SIR_model[,3], col="cornflowerblue", lwd=2)

```


**Excercises**

- 

-  

